GEE models

### 2017-07-20 doc/manuscript.Rmd
reformat paragraphs
multiple clarifications for stats analysis in methods

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA2-%S largely reflects hepatic insulin sensitivity, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Hermans1999a; @Matsuda1999; @Retnakaran2009]. 
Formulas used to calculate these measures can be found in the ESM Methods.

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations (see `r cite_sf("dag")`), and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC. To reduce
complexity between models using either insulin sensitivity or beta-cell function
measures, only the model that had similar good fit for both was used. 
As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures (M9). The final GEE model was adjusted for years from baseline (time),
sex (female or male), ethnicity (European ancestry or non-European ancestry),
baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, and baseline 
age were time-*independent* variables. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every
standard deviation (SD) increase in the predictor variable given the covariates
are held constant.
Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
production, TAG may act as a mediator between NEFA and the outcomes. To
determine the role of TAG in the association between NEFA and the outcomes, TAG
was included in the GEE model as a sensitivity analysis. Lastly for the GEE
models, we tested for an interaction with sex, ethnicity, or time by the
predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. Therefore, to confirm the results of the GEE
analyses in a multivariate environment (all NEFA variables in a single model),
partial least squares (PLS) was used with the outcomes as the independent variable.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a].
The R code for this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered statistically significance at p<0.05,
after adjusting for multiple testing using the Benjamini-Hochberg False
Discovery Rate [@Benjamini1995a]. STROBE was used as a guideline for reporting [@Vandenbroucke2007b].

>>>>

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  Beta-cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle missed visits. The predictor variables and continuous covariates were
scaled (mean centered and standardized). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC). See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. After scaling, 
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable given the covariates are held constant. With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction, as each model only contains data from a single
fatty acid. Therefore, to confirm the results of the GEE analyses in a
multivariate environment (all NEFA variables in a single model), partial least
squares (PLS) was used with the outcomes as the dependent variable.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].
